Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2024-08-22 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

@@

Given observations
\(\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)\), we would
like to compute the posterior distribution over the random variables
\begin{align}
\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align} but exact Bayesian inference is intractable in this model.
We use Pyro to automatically integrate out the local discrete latent
variables \(k\), which is defined as the cell and gene transcriptional
state (see above), and approximate the posterior over the remaining
latent variables using variational inference
\citep{Bingham2018-id, Kucukelbir2016-bk}, which converts intractable
integrals into optimization problems that can be solved with
off-the-shelf tools. In variational inference, we maximize a tractable
bound (the evidence lower bound (ELBO)) on the Kullback-Leibler (KL)
divergence
\[\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} p(\theta, \psi \vert \tilde{X}_{cg}) \right)\]
between a parametric family of tractable probability distributions q and
the intractable true posterior \begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align} We approximate our model's posterior distribution
\(p( \theta, \psi \vert
\tilde{X})\) with a tractable family of probability distributions
\(q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)\), where
\(q_{\phi_1}(\theta)\) is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
\(q_{\phi_2}(\psi)\) is a multivariate Gaussian distribution with a
learnable location parameter and a learnable low-rank covariance matrix.
We solve the resulting stochastic optimization problem using a version
of Automatic Differentiation Variational Inference (ADVI)
\citep{Kucukelbir2016-bk}: we obtain gradient estimates by
differentiating a Monte Carlo estimate of the ELBO and update the
variational parameters using stochastic gradient ascent. Our
implementation and experiments use generic variational families and
Monte Carlo ELBO and gradient estimators provided by Pyro
\citep{Bingham2018-id}, an open source library for probabilistic machine
learning, and a version built into Pyro of the adaptive stochastic
gradient ascent algorithm Adam augmented with gradient clipping to
enhance numerical stability during training.

@@

Pyro -Velocity provides two model formulations as described above. Users
can compare their predictive performance to decide which of the two
models to use for a given dataset. Namely, the mean absolute errors of
posterior samples prediction of spliced and unspliced raw read counts
\(x_{icg}\) from its observation \(\tilde{X}_{cg}\) per gene and cell
combinations \begin{align*}
\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} |x_{icg} - \tilde{X}_{cg}|,
\end{align*} where \(i\) is the posterior sample index, \(N_i\) is the
posterior sample number, \(N_c\) is the cell number, \(N_g\) is the gene
number. Then it is possible to evaluate the cell fate inference
performance of the models based on two metrics:

@@

\caption{\label{extfig-lineage-tracing}Extension of
Fig.~\ref{fig-lineage-tracing}, Pyro -Velocity (Model 1 for uni-fate
data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
scRNA-seq datasets with lineage barcoding information (LARRY dataset).
\textbf{a}, SPRING embedding rendering of Cytotrace scores, the title
reports the Spearman's correlation between the Cytotrace and Cospar fate
potency scores, \textbf{b}, Pyro -Velocity additional single cell level
uncertainty diagnostic plots based on the multi-fate LARRY dataset, from
left to right: 1. Standard deviation of posterior samples of shared time
per cell, 2. Rayleigh test of posterior samples of vector field per cell
(the title reports the false discovery rate after p-value correction for
multiple tests across cells), 3. Transcriptional state uncertainty of
predicted raw read count per cell. The contour lines show the top 10\%
of the most uncertain cells (\hyperref[sec-methods]{Methods}),
\textbf{c}, Pairwise Spearman's correlation between Cytotrace/Cospar
fate potency scores and velocity methods for the four different
benchmarking datasets selected based on the lineage barcoding
information from LARRY data, \textbf{d}, SPRING embedding rendering of
posterior mean of shared time across 30 posterior samples, the title
shows the Spearman's correlation of mean shared time with Cytotrace.
\textbf{e}, Selection of the top positively correlated genes with the
cell shared time based on the posterior mean of spliced expression.
\textbf{f}, Phase portraits, rainbow plots, and UMAP rendering with
splicing gene expression level for the selected genes in \textbf{e}.}

>>>>

We use these solutions to formulate an end-to-end probabilistic
generative model that relates prior distributions on kinetic parameters
to a distribution on pairs of observed unspliced and spliced read count
matrices

\begin{align}
\alpha^{(0)}{ }_g &\sim \operatorname{LogNormal}(0,1), \\
\beta_g &\sim \operatorname{LogNormal}(0,1), \\
\gamma_g &\sim \operatorname{LogNormal}(0,1), \\
&\hskip -18pt \Delta \text { switching }_g \sim \operatorname{LogNormal}(0,1), \\
t_0^{\left(k_{c g}\right)} &= \left\{
  \begin{array}{l}
    t_0^{(0)}{ }_g \sim \operatorname{Normal}(0,1), k_{c g}=0 \\
    t_0^{(1)}{ }_g=t_0^{(0)}{ }_g+\Delta \text { switching }_g, \\
    \quad k_{c g}=1
  \end{array}\right. \\
t_c &\sim \operatorname{LogNormal}(0,1), \\
k_{c g} &\sim \text{Bernoulli} \left( \text{logits}= t_c-t_0^{(1)} \right), \\
\tau^{\left(k_{c g}\right)} 
  &= \operatorname{softplus}\left(t_c-t_0^{\left(k_{c g}\right)}{ }_g\right), \\
u_{c g} 
 &= \text { Measurement }_u \left( u\left(\tau^{\left(k_{c g}\right)}\right) ; 
                                   u_{c g}^{obs}\right), \\
s_{c g} 
  &= \text { Measurement }_s \left( s\left(\tau^{(k_{c g})}\right) ; 
                                    s_{c g}^{obs}\right).
\end{align} \(u\left(\tau^{\left(k_{c g}\right)}\right)\) and
\(s\left(\tau^{(k_{c g})}\right)\) are are called the denoised gene
expression calculated from the velocity analytic solution input with the
kinetics random variables. \(u_{cg}\) and \(s_{cg}\) are the spliced and
unspliced read count sampled from the Poisson models. \(u_{cg}^{obs}\)
and \(s_{cg}^{obs}\) are the observed spliced and unspliced read count
tables. The generative process

\(\text{Measurement}(\cdot)\) for observed unspliced read counts given
denoised unspliced gene transcript expression level
\(u\left(\tau^{(k_{cg})}\right)\) (and identical for observed spliced
read counts) models the expected number of observed reads for a given
gene in a given cell as the number of transcripts times the ratio of the
cell's total reads to total transcripts \begin{align}
u_c^{\hat{obs}} &= \sum_g u_{c g}^{obs}, \\
\hat{u}_c &= \sum_g u\left( \tau^{(k_{c g})}\right), \\
\eta_c^{(u)} &\sim \operatorname{Normal}\left(
    u_c^{\hat{obs}_c}, 
    \operatorname{std} \left(u_c^{\hat{obs}}\right)
  \right), \\
\mu_{c g}^{(u)} &= \log \left(u\left(\tau^k{ }_{c g}\right)\right)
  +\log \left(\eta_c^{(u)}\right)-\log \left(\hat{u}_c\right), \\
u_{c g}^{obs} &\sim 
  \operatorname{Poisson}\left(\lambda=\exp \left(\mu_{c g}^{(u)}\right)\right).
\end{align}

For the first Pyro-Velocity model (Model 1), we constrain the shared
time to be strictly larger than \(t_{0}^{(0)}\) by introducing auxiliary
random variables \[
\text{t\_constraint}_{cg} 
  \sim \text{Bernoulli} \left( \text{logits} = t_c - {t_{0}^{(0)}}_g \right),
\] and setting their values to \(1\), and we set the initial condition
per gene to be \begin{align}
\left( {u_{0}^{(k_{cg})}}_g , {s_{0}^{(k_{cg})}}_g \right) &= \left\{
  \begin{array}{l}
    (0,0), k_{c g}=0 \\
    \bigg( {u \left( \Delta \text { switching }_g \right)}_g,\\
           \quad {s \left( \Delta \text { switching }_g \right)}_g \bigg), \\
           \quad k_{c g}=1
  \end{array}\right.
\end{align} For the extended Pyro -Velocity model (Model 2), we remove
the shared time constraint \(\text{t\_constraint}_{cg}\), thus allowing
a time lag per gene that might be caused by delayed gene activation and
set the initial condition per gene as random variables that are strictly
positive \(\left({u_{0}^{(0)}}_g,
{s_{0}^{(0)}}_g\right)\), which allow genes having a basal expression
level before gene activation. Then, we compute the gene expression at
the second switching time as \begin{align}
({u_{0}^{(1)}}_g, {s_{0}^{(1)}}_g) &= 
  \bigg( {u \left( \Delta \text { switching }_g \right)}_g, \nonumber \\
& \qquad {s \left( \Delta \text { switching }_g \right)}_g \bigg),
\end{align} which shares the same initial condition
\(\left({u_{0}^{(0)}}_g, {s_{0}^{(0)}}_g\right)\) where \begin{align}
{u_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1),\\
{s_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1).
\end{align}

@@

Given observations
\(\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)\), we would
like to compute the posterior distribution over the random variables
\begin{align}
\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, 
          {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align} but exact Bayesian inference is intractable in this model.
We use Pyro to automatically integrate out the local discrete latent
variables \(k\), which is defined as the cell and gene transcriptional
state (see above), and approximate the posterior over the remaining
latent variables using variational inference
\citep{Bingham2018-id, Kucukelbir2016-bk}, which converts intractable
integrals into optimization problems that can be solved with
off-the-shelf tools. In variational inference, we maximize a tractable
bound (the evidence lower bound (ELBO)) on the Kullback-Leibler (KL)
divergence \[
\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} 
  p(\theta, \psi \vert \tilde{X}_{cg}) \right)
\] between a parametric family of tractable probability distributions q
and the intractable true posterior \begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} 
     \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align} We approximate our model's posterior distribution
\(p( \theta, \psi \vert
\tilde{X})\) with a tractable family of probability distributions
\(q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)\), where
\(q_{\phi_1}(\theta)\) is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
\(q_{\phi_2}(\psi)\) is a multivariate Gaussian distribution with a
learnable location parameter and a learnable low-rank covariance matrix.
We solve the resulting stochastic optimization problem using a version
of Automatic Differentiation Variational Inference (ADVI)
\citep{Kucukelbir2016-bk}: we obtain gradient estimates by
differentiating a Monte Carlo estimate of the ELBO and update the
variational parameters using stochastic gradient ascent. Our
implementation and experiments use generic variational families and
Monte Carlo ELBO and gradient estimators provided by Pyro
\citep{Bingham2018-id}, an open source library for probabilistic machine
learning, and a version built into Pyro of the adaptive stochastic
gradient ascent algorithm Adam augmented with gradient clipping to
enhance numerical stability during training.

@@

Pyro -Velocity provides two model formulations as described above. Users
can compare their predictive performance to decide which of the two
models to use for a given dataset. Namely, the mean absolute errors of
posterior samples prediction of spliced and unspliced raw read counts
\(x_{icg}\) from its observation \(\tilde{X}_{cg}\) per gene and cell
combinations \begin{align*}
\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} 
  \vert x_{icg} - \tilde{X}_{cg} \vert,
\end{align*} where \(i\) is the posterior sample index, \(N_i\) is the
posterior sample number, \(N_c\) is the cell number, \(N_g\) is the gene
number. Then it is possible to evaluate the cell fate inference
performance of the models based on two metrics:

@@

\caption{\label{extfig-lineage-tracing}Extension of
Fig.~\ref{fig-lineage-tracing}, Pyro -Velocity (Model 1 for uni-fate
data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
scRNA-seq datasets with lineage barcoding information (LARRY dataset).
\textbf{a}, SPRING embedding rendering of Cytotrace scores, the title
reports the Spearman's correlation between the Cytotrace and Cospar fate
potency scores, \textbf{b}, Pyro -Velocity additional single cell level
uncertainty diagnostic plots based on the multi-fate LARRY dataset, from
left to right: 1. Standard deviation of posterior samples of shared time
per cell, 2. Rayleigh test of posterior samples of vector field per cell
(the title reports the false discovery rate after p-value correction for
multiple tests across cells), 3. Transcriptional state uncertainty of
predicted raw read count per cell. The contour lines show the top 10\%
of the most uncertain cells (\hyperref[sec-methods]{Methods}),
\textbf{c}, Pairwise Spearman's correlation between Cytotrace/Cospar
fate potency scores and velocity methods for the four different
benchmarking datasets selected based on the lineage barcoding
information from LARRY data, \textbf{d}, SPRING embedding rendering of
posterior mean of shared time across 30 posterior samples, the title
shows the Spearman's correlation of mean shared time with Cytotrace.
\textbf{e}, Selection of the top positively correlated genes with the
cell shared time based on the posterior mean of spliced expression.
\textbf{f}, Phase portraits, rainbow plots, and UMAP rendering with
splicing gene expression level for the genes selected in \textbf{e}.}
